Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-21 main.tex
Completed a pass of review and editing, manually adjusted some references.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range between 50\% and 95\% in other surveyed
countries~\cite{Wegman2024}. Utriainen, O’Hern, and
Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
control, and engagement with a curb, pothole, or bollard are most the common
fall types in the seven reported countries. Most of these types involve
perturbations that result in loss of balance. At low speeds, from start up to
typical cruising speeds, bicycles are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of some of their necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
known that applying a motor actuated steering torque proportional to the
vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
J670 vehicle standards.} can stabilize a single-track vehicle down to very low
speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \cite{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\cite{Kane1985}. We chose to perturb the
handlebar because of this superposition principle and that it was both
practical and required a low force to cause a fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
Whipple~\cite{Whipple1899} not only derived the correct equations of motion of
the bicycle \cite{Meijaard2007} but realized and showed that roll motion
feedback can stabilize a bicycle. Much later, attempts at automatic roll
stabilization of single-track vehicles began in earnest after predictive
motorcycle models were developed and refined throughout the 1970s. Van
Zytveld~\cite{vanZytveld1975} was influenced by Whipple's work and seems to be
the first to attempt to robotically stabilize a small motorbike with a
controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that Nagai~\cite{Nagai1983} successfully demonstrated balancing a
robotic bicycle on a treadmill with both steering control and a laterally
moving mass. Ruijs and Pacejka~\cite{Ruijs1986} followed this breakthrough by
demonstrating an automatically balanced motorcycle and they did so solely with
a steering motor. Ruijs and Pacejka showed that steer torque driven by roll
angle feedback stabilizes the capsize mode, by roll angular rate feedback
stabilizes the weave mode, and by steer angular rate feedback stabilizes the
wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes are defined in
\cite{Sharp1971}.} They also showed how the control gains must change with
respect to vehicle speed for favorable control across all speeds.  Thus roll
motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

@@

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
but the time constant of the unstable eigenvalue is increased by a factor of 8
and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
also shows the identified weave mode captured using the methods of
Kooijman~et~al.~\cite{Kooijman2009a} and gives an indication of the performance
of the system, with the model predicting the weave mode well the region of the
tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and the simulate
real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
characterize the configuration of the bicycle when perturbed. The sense of
these variables relative to the sense of the perturbation matters. For example,
if there is already a large steer angle in the same direction as the
perturbation, this will have a different effect on the probability that a fall
will occur than if the steer angle is in the opposite direction of the
perturbation. To account for this, the roll and steer variables at the instance
of perturbation are normalized for perturbation direction. The gain factor
setting on the balance assist controller indicates if the assistance is off
\(\kappa=0\) or on at two different levels: low \(\kappa=3.9\) or high
\(\kappa=5.2\). As mentioned earlier, a recovery from the perturbation is
successful if the person neither places their foot down onto the treadmill
surface nor their wheel of the bicycle exits the width of the treadmill belt.
We record this as a binary variable \(f\) for ``fall''.  All measured variables
are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Due to data quality issues that prevented time synchronization of the
measurements, we excluded one participant from the 6~\si{\kps} dataset and one
from the 10~\si{\kps} dataset, leaving 10 and 14 included participants with 40
perturbations each. Before fitting the model, we scale each independent
variable \(x_{ij}^k\) such that they have a mean of zero and a standard
deviation of one by cluster-mean centering with clusters being an individual
participant, as recommended by \cite{Enders2007}. The clusters are chosen as
all data from an individual participant because we are only interested in the
association between the state of the balance assist system and the outcome of
the perturbation for that participant, not how participants perform relative to
each other. We expected there to be variation between participants in how well
they are able to resist the perturbation; however, this was not true. If a
random effect is included in the model for each participant, the intraclass
correlation (ICC) is less that 3\% for both the 6~\si{\kph} and 10~\si{\kph}
datasets, so we utilize a simple single-level logistic regression model without
the random effect instead of a multilevel model. This tracks with the
independent data collected in \cite{Haitjema2023a} which also had negligible
participant variation. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
for each set of data.  Statistics were computed with R version
4.1.3~\cite{RCoreTeam2021} the lme4 version 1.1~\cite{Bates2015} package.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=5.2\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
to calculate the probability of falling.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) / \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) / \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain factor \(\kappa=3.9\) are shown in Table
\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
statistically significant predictors with angular impulse being the dominant
effect. Larger angular impulse increases the probability to fall and enduring
more perturbations decreases the probability to fall. Unlike the 6~\si{\kph}
trials, the balance assist state is not a significant predictor (\(p=0.07\))
but the effect would be a reduction in the probability to fall if it were. At
both speeds, the angular impulse has about twice the magnitude in effect as the
perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) / \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) / \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(s\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range between 50\% and 95\% in other surveyed
countries~\cite{Wegman2024}. Utriainen, O'Hern, and
Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
control, and engagement with a curb, pothole, or bollard are the most common
fall types in the seven reported countries. Most of these types involve
perturbations that result in loss of balance. At low speeds, from start up to
typical cruising speeds, bicycles are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of some of their necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
known that applying a motor actuated steering torque proportional to the
vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
J670 vehicle standards.} can stabilize a single-track vehicle down to very low
speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack; first introduced in \cite{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not sufficient to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\cite{Kane1985}. We chose to perturb the
handlebar because of this superposition principle, it was practical, and it
required a low force to cause a fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
Whipple~\cite{Whipple1899} not only derived the correct equations of motion of
the bicycle~\cite{Meijaard2007} but realized and showed that roll motion
feedback can stabilize a bicycle. Much later, attempts at automatic roll
stabilization of single-track vehicles began in earnest after predictive
motorcycle models were developed and refined throughout the 1970s. Van
Zytveld~\cite{vanZytveld1975} was influenced by Whipple's work and seems to be
the first to attempt to robotically stabilize a small motorbike with a
controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that Nagai~\cite{Nagai1983} successfully demonstrated balancing a
robotic bicycle on a treadmill with both steering control and a laterally
moving mass. Ruijs and Pacejka~\cite{Ruijs1986} followed this breakthrough by
demonstrating an automatically balanced motorcycle and they did so solely with
a steering motor. Ruijs and Pacejka showed that steer torque driven by roll
angle feedback stabilizes the capsize mode, by roll angular rate feedback
stabilizes the weave mode, and by steer angular rate feedback stabilizes the
wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes are defined in
\cite{Sharp1971}.} They also showed how the control gains must change with
respect to vehicle speed for favorable control across all speeds.  Thus roll
motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

@@

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Notably, the controlled bicycle is not stable at
6~\si{\kph} but the time constant of the unstable eigenvalue is increased by a
factor of eight and at 10~\si{\kph} the system is just stable.
Figure~\ref{fig:eigenvalues} also shows the identified weave mode captured
using the methods of Kooijman~et~al.~\cite{Kooijman2009a} and gives an
indication of the performance of the system, with the model predicting the
weave mode well the region of the tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and they simulate
real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
characterize the configuration of the bicycle when perturbed. The sense of
these variables relative to the sense of the perturbation matters. For example,
if there is already a large steer angle in the same direction as the
perturbation, this will have a different effect on the probability that a fall
will occur than if the steer angle is in the opposite direction of the
perturbation. To account for this, the roll and steer variables at the instance
of perturbation are normalized for perturbation direction. The gain factor
setting on the balance assist controller indicates if the assistance is off
\(\kappa=0\) or on at two different levels: low \(\kappa=3.9\) or high
\(\kappa=5.2\). As mentioned earlier, a recovery from the perturbation is
successful if the person neither places their foot down onto the treadmill
surface nor the wheel of the bicycle exits the width of the treadmill belt.  We
record this as a binary variable \(f\) for ``fall''.  All measured variables
are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Due to data quality issues that prevented time synchronization of the
measurements, we excluded one participant from the 6~\si{\kph} dataset and one
from the 10~\si{\kph} dataset, leaving 10 and 14 included participants with 40
perturbations each. Before fitting the model, we scale each independent
variable \(x_{ij}^k\) such that they have a mean of zero and a standard
deviation of one by cluster-mean centering with clusters being an individual
participant, as recommended by \cite{Enders2007}. The clusters are chosen as
all data from an individual participant because we are only interested in the
association between the state of the balance assist system and the outcome of
the perturbation for that participant, not how participants perform relative to
each other. We expected there to be variation between participants in how well
they are able to resist the perturbation; however, this was not true. If a
random effect is included in the model for each participant, the intraclass
correlation (ICC) is less that 3\% for both the 6~\si{\kph} and 10~\si{\kph}
datasets, so we utilize a simple single-level logistic regression model without
the random effect instead of a multilevel model. This tracks with the
independent data collected in \cite{Haitjema2023a} which also had negligible
participant variation. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
for each set of data.  Statistics were computed with R version
4.1.3~\cite{RCoreTeam2021} the lme4 version 1.1~\cite{Bates2015} package.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor \(\kappa=5.2\) trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
to calculate the probability of falling.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain factor \(\kappa=3.9\) are shown in Table
\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
statistically significant predictors with angular impulse being the dominant
effect. Larger angular impulse increases the probability to fall and enduring
more perturbations decreases the probability to fall. Unlike the 6~\si{\kph}
trials, the balance assist state is not a significant predictor (\(p=0.07\))
for our \(p\)-value threshold but the effect would be a reduction in the
probability to fall if it were. At both speeds, the angular impulse has about
twice the magnitude in effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
    standard error SE, p-value \(p\), multiplicative change in odds
    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(s\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}
